Self Referencing Sequences

### 2014-11-07 draft.tex
edited rules section

Why does this work? Well, first of all, notice that the rules are the natural applications of the self-referential sequence. Also, note that we have decided to make the rules of length 2; if we had made the self-referential rules $1 \rightarrow 1$ and $3 \rightarrow 333$, this would not have worked. In addition, we know that each block length preserves parity.

@@

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating cycle to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.

Therefore, we can formulate a set of sufficient conditions to make rules to generate the self-referential sequence with generating cycle $C$.

We want to have a set of rules $R$ where each rule $r \in R$ has length $|r| = l$, such that all $|C|^{l}$ strings are in the rule set and for each $r \in R$
\begin{itemize}
\item each rule writes blocks in the same order of $C$ and each rule has the same starting number
\item $\sum_{i \in r} i$ is divisible by $l$
\item $l$ is divisible by $|C|$
\end{itemize}
Given these conditions on every rule, we can guarantee that the self-referential sequence is generated. Note that in this case, the above example does not satisfy our rules until we say that $1, 3 \to 3, 1, 1, 1$.

Another example, we can have $C = (2 4)$, and we can have that the rules be
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}

One of the reasons why generating a rule set for $(1 2)$ is complicated is that since $1, ..., 1$ must be in the rule set, the length of the rules must be even (since $|C| = 2$). On the other hand, $2, 1, ..., 1$ must also be in the set and so the rule length must be odd.

>>>>

\begin{comment}
Why does this work? Well, first of all, notice that each rule of the form $x \rightarrow y$ has the . Also, note that we have decided to make the rules of length 2; if we had made the self-referential rules $1 \rightarrow 1$ and $3 \rightarrow 333$, this would not have worked. In addition, we know that each block length preserves parity.

@@

When we apply our substitution rules we should not need to know what happened to blocks other than the one we are currently substituting for. Therefore, each rule must somehow encode the length of  blocks, the number of elements already written in the current block, and what number the current block is composed of. The first condition is simple to implement, as the rules will specify the length of blocks. In order to keep track of the numbers already written, each rule should be self-contained. That is each rule should write complete blocks. In order to comply with the third condition it is enough for the size of the generating cycle to divide the length of the rules. This is so that at the start of each rule, we always know which numbers to write.
\end{comment}

We can formulate a set of sufficient conditions for rules that generate the self-referencing sequences. Formally, we will set $R$ to be a rule set, which is a set of strings of the form $x \rightarrow y$.

\begin{proposition}
\label{ruleprop}
Suppose $C$ is a cycle. For any $l \in \mathbb{N}$ divisible by $|C|$ and $a \in C$, let $R$ be a rule set. If for each string $s$ of elements in $C$ of length $l$, there exists a unique rule $x \rightarrow y$ in $R$ with $x = s$ and:
\begin{enumerate}
\item $x$ reads the block lengths of $y$ from left to right
\item $y$ starts with $a$
\item $|y|$ is divisible by $l$
\end{enumerate}
Then sucessively applying the rules in $R$, starting with a sequence of length $l$ generates a sequence which is weakly convergent to $s(C, a)$.
\end{proposition}

\begin{proof}
Let $s^{(i)}$ be the sequence generated from the process from Section~\ref{iterativeprocess}. I will show by induction on $i$ that $s^{(i)}$ is the result of applying the rules in $R$ $i$ times to a starting sequence $s^{(0)}$ of length $l$. By letting the starting sequence be $s^{(0)}$, we know it works for $i = 0$.

Suppose that applying the rules in $R$ to $s^{(0)}$ $i$ times gives $s^{(i)}$. Note that $|s^{(i)}|$ is divisible by $l$ since each iteration applies the rule to a segment of length $l$ and by condition 3, it writes a sequence of length a multiple of $l$. So we can partition $s^{(i)}$ uniquely into disjoint chunks of size $l$.

In the iterative process, each chunks of size $l$ will generate $l$ blocks, and since $|C|$ divides $l$, the block referenced by each of these chunks will have the same number. If the rules begin with $a$, then this number will be $a$.

So applying the rules to each chunk is equivalent to applying the iterative process, where we  group elements into chunks of size $l$. Therefore, $s^{(i+1)}$ is the result of applying the rules to $s^{(i)}$.
\end{proof}

Given Proposition~\ref{ruleprop} we can generate rules a large class of generating cycles $C$. Take the following example, where $C = (2 4)$, and the rules are
\begin{align*}
\text{Rule }A: 2,2 &\to 2, 2, 4, 4 \\
\text{Rule }B: 2,4 &\to 2, 2, 4,4,4,4\\
\text{Rule }C: 4,2 &\to 2,2,2,2,4,4\\
\text{Rule }D: 4,4 &\to 2, 2, 2, 2, 4, 4, 4, 4
\end{align*}
These rules satisfy the conditions of Proposition~\ref{ruleprop}, and so successive application of these rules will generate a self-referencing sequence which is weakly convergent to $s(C, 2)$.

\begin{proposition}
\label{norules12}
There is no set of rules $R$ satisfying the conditions of Proposition~\ref{ruleprop} for $C=(12)$.
\end{proposition}

\begin{proof}
Suppose $l$ is a positive integer. Then $l$ must be even. Let $s$ be the string with one 2 and $l-1$ 1's. Suppose $x \rightarrow y$ is the rule for $s$, so $x = s$. Then by condition 1 of Proposition~\ref{ruleprop}, $|y| = l+1$, and by condition 3, $l = 1$. Which contradicts the fact that $l$ is divisible by $|C| = 2$. 
\end{proof}
